% =====================================================================
%  Communications AI & Computing (Nature Portfolio) -- submission draft
%  Built from the VERBATIM TPAMI-revised section sources, reordered to the
%  Nature house structure (Methods at the end), with surgical additions for
%  the collagen sponge dataset only. Original text and voice are preserved.
% =====================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathptmx}                 % Times-like text + math
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{siunitx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{authblk}
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}
\usepackage[font=small,labelfont=bf]{caption}

\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds, calc, patterns}
\pgfplotsset{compat=1.18}

% --- Nature-style numeric superscript citations ---
\usepackage[numbers,super,sort&compress]{natbib}

% --- Custom macros used by the original section sources ---
\newcommand{\lambdaex}{\lambda_{\text{ex}}}
\newcommand{\lambdaem}{\lambda_{\text{em}}}
\providecommand{\R}{}\renewcommand{\R}{\mathbb{R}}
\newcommand{\vect}[1]{\mathbf{#1}}
\newcommand{\mat}[1]{\mathbf{#1}}

% --- Shim for the IEEEtran drop-cap macro used in the introduction ---
\providecommand{\IEEEPARstart}[2]{#1#2}

\doublespacing
\linenumbers

% Slightly relax float restrictions for the single-column layout
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}
\renewcommand{\textfraction}{0.05}
\renewcommand{\floatpagefraction}{0.85}

\title{\bfseries Unsupervised deep learning for wavelength selection in multi-excitation fluorescence imaging}

\author[1,2]{Narek Meloyan}
\author[1,2,3]{Narine Sarvazyan}
\affil[1]{American University of Armenia, Yerevan, Armenia}
\affil[2]{L.~A.~Orbeli Institute of Physiology NAS RA, Yerevan, Armenia}
\affil[3]{George Washington University, Washington, DC, USA}
\date{}

\begin{document}
\maketitle

% ---- Abstract (original text + one collagen sentence) ----
\input{sections/abstract}

\vspace{0.5em}
\noindent\textbf{Keywords:} wavelength selection, convolutional autoencoder,
perturbation analysis, spectral band selection, fluorescence imaging.

% =====================================================================
%  NARRATIVE (Nature order: Introduction -> Results -> Discussion)
%  Section text is the original, verbatim, with surgical collagen edits.
% =====================================================================
\input{sections/introduction}
\input{sections/related_work}
\input{sections/results}
\input{sections/discussion}
\input{sections/conclusion}

% =====================================================================
%  METHODS (placed at the end per Nature Portfolio convention).
%  These are the original Methodology and Experimental Setup sections.
% =====================================================================
\input{sections/methodology}
\input{sections/experimental_setup}

% =====================================================================
%  BACK MATTER
% =====================================================================
\section*{Data availability}
The datasets analysed in this study are available from the corresponding author
on reasonable request.

\section*{Code availability}
The implementation is available at \url{https://github.com/Nard248/spectral-select}.

\section*{Acknowledgements}
Financial support of the European Union (ERA Chair NAS-SAR award, grant
101087403) is gratefully acknowledged. The authors thank Dr.~Arsen Gasparyan
(Takhtajyan Institute of Botany, National Academy of Sciences of Armenia) for
providing lichen samples, Narek Chillingaryan (L.~A.~Orbeli Institute of
Physiology) for assistance with sample preparation and data acquisition, and
Aleksandr Hayrapetyan (Yerevan State University) for assistance with model
selection and benchmarking. A large-language-model assistant was used solely to
refine grammatical clarity and to edit source files; it did not contribute to
the conceptual design, data analysis, or intellectual content of this work.

\section*{Author contributions}
N.M. designed the framework, implemented the experiments, and drafted the
manuscript. N.S. supervised the study, provided the imaging data and domain
interpretation, and revised the manuscript. Both authors approved the final
version.

\section*{Competing interests}
The authors declare no competing interests.

% =====================================================================
\bibliographystyle{naturemag}
\bibliography{references}

\end{document}
